Interisland movements of woodland caribou in Newfoundland

### 2020-02-19 paper/manuscript.Rmd
edits reflow yes

In Newfoundland, wolves (Canis lupus) are extirpated, so coyotes (Canis latrans)
and black bears (Ursus americanus) are the primary predators of caribou
(Bastille-Rousseau et al. 2016). Coyotes and black bears are responsible for the
majority of mortalities for neonate caribou calves (Bastille-Rousseau et al.
2016), although predation can still occur after the calving period (Lewis &
Mahoney 2014). However, on Fogo Island, coyotes are the only predator of
caribou.

Forage limitation could explain movement between islands. Forage scarcity has
been proposed as a potential reason why caribou move between arctic islands on
the sea-ice in winter (Miller et al. 1977). For very small islands, over-grazing
by newly arrived caribou may rapidly limit forage (Bergerud et al. 1990).
Without sea ice, movement from the mainland to an island as well as between
islands requires caribou to assess the trade-off associated with swimming. The
underlying mechanism driving forage limitation and the trade-off associated with
swimming could be density-dependent habitat selection governed by the Ideal Free
Distribution (Morris 1987; Bradbury et al. 2015). Ideal Free Distribution theory
suggests that habitat selection is density-dependent and that variation in
density between habitat patches leads to a fitness equilibrium (Bradbury et al.
2015). Fitness in a habitat patch depends on density where, ideally, the
available resources on a habitat patch can sustain a specific number of
individuals. In a hypothetical example, when density of a given habitat patch
has exceeded the fitness equilibrium, animals are predicted to relocate and
settle new habitat patches to reach equilibrium. Islands therefore represent
distinct habitat patches with clear costs and benefits associated with swimming.

Woodland caribou are well known to disperse to remote locations, including
islands, during calving (Bergerud & Page 1987; Cumming & Beange 1987). Moving to
remote locations is a viable anti-predator strategy for caribou because it
reduces detection and encounter rates by predators. For instance, the use of
shoreline habitat in summer was deemed an effective anti-predator calving
strategy because caribou avoided predators by using islands and peninsulas or
swimming as a flight response (Bergerud et al. 1990). The use of water, and by
association occasionally swimming, is therefore part of the fine-scale
interactions between caribou and their predators and the use of islands appears
to be an effective anti-predator strategy.

Forage limitation and over-grazing is a major issue for caribou populations
(Schaefer et al. 2016; Zamin et al. 2017). Reduced access to foraging
opportunities can result in higher conspecific competition for resources and
lowered reproductive success for adult females (Schaefer et al. 2016). For
caribou in the Fogo Island archipelago, the role of forage limitation in animal
habitat selection patterns and movement remains unknown, but it is possible that
forage limitation and increased conspecific competition influences swimming
between islands.

@@

Predator avoidance could contribute to swimming behaviour of caribou in our
system as well as others (Miller 2002). Unlike elsewhere in the range of
caribou, wolves are extirpated from Newfoundland, including Fogo Island. In
Newfoundland, coyotes are the primary predator of caribou (Bastille-Rousseau et
al. 2016a). While predation by coyotes on adult female caribou is possible, it
is relatively rare on Fogo Island (Webber & Vander Wal unpublished data).
Therefore, any effect of predation on swimming behaviour is likely related to
fine-scale encounters between coyotes and caribou which result in caribou
swimming as a flight response. Alternatively, during calving, females may swim
to small islands to give birth (Bergerud & Page 1987) or swim to islands with a
calf at heel to avoid predation. Swimming with a calf at heel has been observed
for caribou in freshwater lakes on Fogo Island (Webber & Vander Wal unpublished
data), so we anticipate it may also be possible in the ocean. We suggest future
studies assess encounter rates among caribou and coyotes to determine what
proportion of encounters result in swimming behaviour.

>>>>

Forage limitation and over-grazing is a major issue for caribou populations
(Schaefer et al. 2016; Zamin et al. 2017) and could explain movement between
islands. Reduced access to foraging opportunities can result in higher
conspecific competition for resources and lowered reproductive success for adult
females (Schaefer et al. 2016). Forage scarcity has been proposed as a potential
reason why caribou move between arctic islands on the sea-ice in winter (Miller
et al. 1977).  For very small islands, over-grazing by newly arrived caribou may
rapidly limit forage (Bergerud et al. 1990). Without sea ice, movement from the
mainland to an island as well as between islands requires caribou to assess the
trade-off associated with swimming.

The underlying mechanism driving forage limitation and the trade-off associated
with swimming could be density-dependent habitat selection governed by the Ideal
Free Distribution (Morris 1987; Bradbury et al. 2015). Ideal Free Distribution
theory suggests that habitat selection is density-dependent and that variation
in density between habitat patches leads to a fitness equilibrium (Bradbury et
al. 2015). Fitness in a habitat patch depends on density where, ideally, the
available resources on a habitat patch can sustain a specific number of
individuals. In a hypothetical example, when density of a given habitat patch
has exceeded the fitness equilibrium, animals are predicted to relocate and
settle new habitat patches to reach equilibrium. Islands therefore represent
distinct habitat patches with clear costs and benefits associated with swimming.
For caribou in the Fogo Island archipelago, the role of forage limitation in
animal habitat selection patterns and movement remains unknown, but it is
possible that forage limitation and increased conspecific competition influences
swimming between islands.

Predator avoidance could contribute to swimming behaviour of caribou (Miller
2002). Unlike elsewhere in the range of caribou, wolves (*Canis lupus*) are
extirpated from Newfoundland, including Fogo Island, and coyotes (*Canis latrans*) and black bears
(*Ursus americanus*) are the primary predators of caribou (Bastille-Rousseau et
al. 2016). While predation by coyotes on adult female caribou is possible, it is
relatively rare on Fogo Island (Webber & Vander Wal unpublished data).
Therefore, any effect of predation on swimming behaviour is likely related to
fine-scale encounters between coyotes and caribou which result in caribou
swimming as a flight response. Alternatively, during calving, females may swim
to small islands to give birth (Bergerud & Page 1987) or swim to islands with a
calf at heel to avoid predation. Swimming with a calf at heel has been observed
for caribou in freshwater lakes on Fogo Island (Webber & Vander Wal unpublished
data), so we anticipate it may also be possible in the ocean. We suggest future
studies assess encounter rates among caribou and coyotes to determine what
proportion of encounters result in swimming behaviour.Woodland caribou are well
known to disperse to remote locations, including islands, during calving
(Bergerud & Page 1987; Cumming & Beange 1987). Moving to remote locations is a
viable anti-predator strategy for caribou because it reduces detection and
encounter rates by predators. For instance, the use of shoreline habitat in
summer was deemed an effective anti-predator calving strategy because caribou
avoided predators by using islands and peninsulas or swimming as a flight
response (Bergerud et al. 1990). The use of water, and by association
occasionally swimming, is therefore part of the fine-scale interactions between
caribou and their predators and the use of islands appears to be an effective
anti-predator strategy. Coyotes and black bears are responsible for the majority
of mortalities for neonate caribou calves (Bastille-Rousseau et al. 2016),
although predation can still occur after the calving period (Lewis & Mahoney
2014). However, on Fogo Island, coyotes are the only predator of caribou.

@@
